---
title: "RaukR_notes"
author: "CPB"
format: html
editor: visual
code-overflow: scroll
affiliation: PMC
date-modified: date-modified
---

# Scripts, functions, and best practices

To run executable Rscript:

1.  Make script executable `chmod +x myscript.R`
2.  First line of script should be `#!/usr/bin/env Rscript`

### Text streams in R (pipe results to script)

```{r}
input_con  <- file("stdin")
open(input_con)
oneline=readLines(input_con, n = 1)
close(input_con)
```

Tidyverse can read a `tibble` from text stream: `read_csv(file("stdin"))`

Any `stdout`produced by the code (`print()`, `cat()`, etc) can be piped to a new process

To write a `tibble`as a text stream: `cat(format_csv(my_tibble))`

### Hidden variables/functions

Hidden functions - for helper functions that the end user doesn't need to access

Still there but users can't see them. Same for hidden variables.

List hidden items `ls(all.names = TRUE)`

### Functions

Inside function - `a <<- "abc"` will assign the variable outside the function - **don't do**

Good practice to give `data` as first argument, especially when using pipes `%>%`

```{r}
#| eval: false
myfun <- function(x, arg)
```

Set arguments to defaults --- better too many args than too few. Also use elipses `…`

#### Closure

Function inside functions. There's some static part to the environment

Double arrow affects global environment one step up, so doesn't affect the "real" global env. So other functions that use `i` are not affected.

```{r}
new_counter <- function() {
  i <- 0
  function() {
    # do something useful, then ...
    i <<- i + 1
    i
  }
}

counter1 <- new_counter(); counter2 <- new_counter()
counter1(); counter1(); counter2()
```

First call to counter1 is `i=1` , counter2 is `i=2`

#### Wrapper functions

With wrapper functions you can set behavior of existing functions/plots. For example, setting the color, size of text, etc

```{r}
my_awesome_plot <- function(x, ...) {
  plot(x, col='red', pch=19, cex.axis=.7, ...)
}
```

### Check

S4 and R6 classes - during OOP block

# Quarto documents

Set `freeze : true` in the yaml header to "cache" the rendered parts. Only the parts you change will be re-rendered. Avoids running and rendering everything again.

```{yaml}
execute:
  freeze: true
```

Chunk options define how chunks behave

-   `eval: false` to not evaluate a code chunk
-   `echo: false` to hide input code
-   `output: true` to show output, `asis` to skip styling
-   `warning: false` hides warnings
-   `message: false` hides messages
-   `error: true` shows error message and continues code execution
-   `include: false` suppresses all output

## Interactive plots

**plotly** - https://plotly.com/ggplot2/ - interactive ggplots with not too much extra work

**rbokeh** - https://hafen.github.io/rbokeh/articles/rbokeh.html

[Interactive stuff documentation](https://quarto.org/docs/interactive/)

## Projects

Common yaml parameters in `_quarto.yml`

```{yaml}
project:
  output-dir: _output

toc: true
number-sections: true
  
format:
  html:
    css: styles.css
  pdf:
    documentclass: report
    margin-left: 30mm
```

[Project documentation](https://quarto.org/docs/projects/quarto-projects.html)

#### Callouts

Show warnings, notes, tips, important, etc

#### Diagrams

Create diagrams with code.

# Debugging

-   Classes and type checking

-   Loops usually fail at the beginning or at the end - testing at the boundaries

-   Give small dataset to test

-   *antibugging* 🕸: `stopifnot(y <= 75)`

## Dumping frames

R state is saved to file, which can be read with a debugger. Allows you to check values of variables

```{r}
options(error = quote(dump.frames(dumpto = "assets/testdump", to.file = T)))
```

## Traceback

Shows what were the function calls and what parameters were passed to them when the error occurred.

## Debug

Using thee `debug` function

```{r}
#| eval: false
h <- function(x, y) { 
  f(x) 
  f(y) 
}

debug(h)
h('text', 7)
undebug(h)
```

## Profiling

### proc.time()

*user time* -- CPU time charged for the execution of user instructions of the calling process,

*system time* -- CPU time charged for execution by the system on behalf of the calling process,

*elapsed time* -- total CPU time elapsed for the currently running R process.

### system.time()

## Optimization

-   vectorize and allocate memory

-   `microbenchmark`

-   Matrix multiplications run faster on GPUs

## Parallelization

-   `parallel`

```{r}
library(parallel)
num_cores <- detectCores() - 1
cl <- makeCluster(num_cores) # Init cluster
parLapply(cl, 1:2, function(x) { c(x, x^2, x^3)} )
stopCluster(cl)
```

# Others

Parenthesis around expression, etc -\> print

```{r}
(vec <- seq(0.1, 0.9, by=0.1))
```

`near` function - safe way of comparing if two vectors of floating point numbers are (pairwise) equal. This is safer than using `==`, because it has a built in tolerance

`simpleWarning` & `simpleErrors`

`ggplot autoplot` - Create a complete ggplot appropriate to a particular data type

# Vectorization

Vectorize user-written functions: `Vectorize` from the base package

```{r}
#| eval: false
vectorized_is_a_droid <- base::Vectorize(is_a_droid, vectorize.args = c('x'))
vectorized_is_a_droid(test)
```

`*apply` takes matrix as input, not vector, need to convert vector to matrix.

`lapply` - apply function to each element of list

`vapply` - you can specify expected output format, error if output not in that format

`purrr` - map function to elements of vector and use `unlist`

# Parallelization

`Future` package

Doesn't block using R during computing. It's possible to keep using R while computing in other cores

> Unblocked R process during resolving of futures process

variable `%<-%` {expression(s)}

-   sequential - use during development

-   multisession -starts different R sessions to parallelize

-   multicore - use different cores for paralleliization

### Sequential

Need to set `plan(sequential)` as first line, like calling `library()`

```{r}
library(future)
plan(sequential)

a %<-% {
  Sys.sleep(3)
  a <- 1
}
b %<-% {
  Sys.sleep(3)
  b <- 2
}

a + b
```

Check number of cores in system `availableCores()`

```{r}
plan(multicore)

a %<-% {
  Sys.sleep(3)
  a <- 1
}
b %<-% {
  Sys.sleep(3)
  b <- 2
}

a + b

```

### Cluster

`plan(cluster)`

```{r}
#| eval: false
plan(cluster, workers = c("n1", "n2", "n3"))
```

Specialized R package for interfacing with common HPC job schedulers exists: `future.batchtools`

# Object oriented programming

Methods inside functions == functions

Classes can inherit properties from other classes

### S4 classes

Object name begins with `.` by convention

`sealed =T` prevents redefining the same class (with same name)

Access slots in class with `@`

# TidyR

### New pipes

-   the `%>%` pipe

-   `x %>% f` ≡ `f(x)`

-   `x %>% f(y)` ≡ `f(x, y)`

-   `x %>% f %>% g %>% h`≡ `h(g(f(x)))`

-   `%T>%` - split data and send to the next two function calls

```{r}
rnorm(50) %>% 
  matrix(ncol = 2) %T>% 
  plot() %>% 
  summary()
```

`%T>%` sends data to both `plot` and `summary`, not `plot` and *then* `summary`

-   `%$%` - for functions that don't take data as the first argument, like cor = `cor(data$1, data$2)`. Exposition of variables.
-   `|>` base R pipe

```{r}
mtcars |> lm(mpg ~ disp, data = _)
```

The `_` placeholder only works for arguments that are named in the function

Sometimes we want to pass the resulting data to *other than the first* argument of the next function in chain. `magritter` provides placeholder mechanism for this:

-   `x %>% f(y, .)` = `f(y, x)`,

-   `x %>% f(y, z = .)` = `f(y, z = x)`.

*\`%\*% (.,.)* - instead of doing x\*y, it will do \*(M,M)

When sending item through pipe, need to use `{}` around the function so that the item only goes to the `.`, so it's not evaluated as the first item in the function

ex: `M %>% {print_M_summ(nrow(.), ncol(.))}`

Functional sequence / sequence of functions: `f <- . %>% sin %>% cos`

## Tibbles

When creating a tibble, a column can be a function of other columns

Change how tibbles are saved

-   `my_tibble %>% print(n = 50, width = Inf)`,
-   `options(tibble.print_min = 15, tibble.print_max = 25)`,
-   `options(dplyr.print_min = Inf)`,
-   `options(tibble.width = Inf)`

#### Parsing

Locale = system definitions, which keyboard, time zone, etc

```{r}
charToRaw(text2)
parse_character(text1, locale = locale(encoding = 'UTF-8'))
guess_encoding(charToRaw("Test"))
guess_encoding(charToRaw(text2))
```

### Data transformations

`arrange` = `order` from data.table

`select(-(x:z))` remove columns x-z

`%>% select(x:z, everything())` reorder columns with x:z in the beginning, and all the rest following

`mutate` create / change variables

`transmute` - only keeps create columns

`group_by` - perform actions per group (columns with grouping)

`n()` count occurrences of category - like data.table `.N`

`pivot_longer` - turn columns into rows

```{r}
bijou2 %>% 
  pivot_longer(cols = c(`2008`, `2009`), names_to = 'year', values_to = 'price') %>% 
  head(n = 5)
```

`pivot_wider` - turn rows into columns

```{r}
bijou3 %>% 
  pivot_wider(names_from = dimension, values_from = measurement) %>% 
  head(n = 5)
```

`unite`  - combine columns

`complete` - complete missing values

`left_join` keep all "keys" from first table

`inner_join` only keep common "keys"

`right_join` keep all keys from second table

`full_join` keep all "keys"

eager - needs the data to exist to use it. lazy - the data can be created in the same table creation call
